---
title: "Midterm"
subtitle: "Predict 413 | Section 59"
author: "Reed Thunstrom"
output: html_document
---

### Problem

Given a set of daily rainfall data, the objective is to predict the next 24 months of precipitation.  The rainfall data is vast, with approximately 66 years of historical data.  These 66 years of daily rainfall data will be used to predicted the next 24 months of rainfall.

### Significance

The problem is an interesting one has much of the world's production comes from rain water.  Farmers no doubt are very interested in monthly rainfall as it provides the fuel for their crops to grow.  Knowing what the precipitation will be is extremely significant and important for them as they need to supply the water themselves if the crops don't get enough water from the rainfall.  

Data: From where did you obtain the data?  What type of data are they?
Type of Models:  What model classes did you build? Why?
Literature:  FIVE examples from PEER REVIEWED JOURNALS of how the types of models you selected were used in similar situations.
Formulation:  How did you formulate / implement you model in R?
Performance / Accuracy: How did your models perform?
Limitations:  What are the limitations of the model you designed?
Future work: Where do you want to go with the model?
Learning:  What did you learn in formulation the model?
Read in Data, create year and month columns.  Aggreate by year and month.  
```{r}
comp_data<- read.csv('/Users/z013nx1/documents/competitiondata.csv', header=TRUE ,sep = ",")
head(comp_data)
comp_data$DATE<- as.Date(as.character(comp_data$DATE), "%Y%m%d")
comp_data["Year"] <- as.numeric(substr(as.character(comp_data$DATE),1,4))
comp_data["Month"] <- as.numeric(substr(as.character(comp_data$DATE),6,7))
comp_data_month_year<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), mean))
df<- data.frame(comp_data_month_year)
comp_data_df<- setNames(df, c("Month", "Year", "PRCP"))
```

Last 6 years.  Multiply per day averages by 30, create time series, remove seasonal effect
```{r}
comp_data_model_6yrs<- comp_data_df[743:815, c("Month", "Year", "PRCP")]
comp_data_model_6yrs$PRCPNEW<- comp_data_model_6yrs$PRCP * 30
monthly_ts<- ts(comp_data_model_6yrs$PRCPNEW, frequency = 12, start = c(2008, 7))
monthly_ts
plot.ts(monthly_ts)
monthly_ts_dc<- decompose(monthly_ts)
plot(monthly_ts_dc)
monthlytsseasadj <- monthly_ts - monthly_ts_dc$seasonal
plot(monthlytsseasadj)
```

Create Holt Winters forecast.  Apply to next 24 months
```{r}
install.packages("forecast", repo="http://cran.us.r-project.org")
library(forecast)
forecasts<- HoltWinters(monthlytsseasadj, beta=FALSE, gamma=FALSE)
forecasts$fitted
plot(forecasts)
forecasts$SSE
next24months<- forecast.HoltWinters(forecasts, h=24)
next24monthsres<- next24months$residuals
next24monthsres<- na.trim(next24monthsres)
```

Examine residuals and autocorrelation
```{r}
acf(next24monthsres, lag.max = 20)
plot.ts(next24monthsres)

plotForecastErrors <- function(forecasterrors)
  {
     # make a histogram of the forecast errors:
     mybinsize <- IQR(forecasterrors)/4
     mysd   <- sd(forecasterrors)
     mymin  <- min(forecasterrors) - mysd*5
     mymax  <- max(forecasterrors) + mysd*3
     # generate normally distributed data with mean 0 and standard deviation mysd
     mynorm <- rnorm(10000, mean=0, sd=mysd)
     mymin2 <- min(mynorm)
     mymax2 <- max(mynorm)
     if (mymin2 < mymin) { mymin <- mymin2 }
     if (mymax2 > mymax) { mymax <- mymax2 }
     # make a red histogram of the forecast errors, with the normally distributed data overlaid:
     mybins <- seq(mymin, mymax, mybinsize)
     hist(forecasterrors, col="red", freq=FALSE, breaks=mybins)
     # freq=FALSE ensures the area under the histogram = 1
     # generate normally distributed data with mean 0 and standard deviation mysd
     myhist <- hist(mynorm, plot=FALSE, breaks=mybins)
     # plot the normal curve as a blue line on top of the histogram of forecast errors:
     points(myhist$mids, myhist$density, type="l", col="blue", lwd=2)
  }
plotForecastErrors(next24monthsres)
plot.forecast(next24months)
next24months
```


Holt Winters with Log transformation

```{r}
plot(comp_data)
summary(comp_data$PRCP)
precipDensity<- density(comp_data$PRCP)
plot(precipDensity)
comp_data$LOG<- log(comp_data$PRCP)
precipLogDensity<- density(comp_data$LOG)
plot(precipLogDensity)
```

```{r}
comp_data_month_year_2<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), sum))
df2<- data.frame(comp_data_month_year_2)
comp_data_df_2<- setNames(df2, c("Month", "Year", "PRCP"))
comp_data_df_2$LOG<- log(comp_data_df_2[,"PRCP"])
```

```{r}
comp_data_model_6yrs_log<- comp_data_df_2[743:815, c("Month", "Year", "LOG")]
comp_data_model_6yrs_log$PRCPNEW<- comp_data_model_6yrs_log$LOG
monthly_ts_log<- ts(comp_data_model_6yrs_log$PRCPNEW, frequency = 12, start = c(2008, 7))
monthly_ts_log
plot.ts(monthly_ts_log)
monthly_ts_log_dc<- decompose(monthly_ts_log)
plot(monthly_ts_log_dc)
monthlytsseasadj_log <- monthly_ts_log - monthly_ts_log_dc$seasonal
plot(monthlytsseasadj_log)
```

Create Holt Winters forecast.  Apply to next 24 months
```{r}
forecasts_log<- HoltWinters(monthlytsseasadj_log, beta=FALSE, gamma=FALSE)
forecasts_log$fitted
plot(forecasts_log)
forecasts_log$SSE
next24months_log<- forecast.HoltWinters(forecasts_log, h=24)
next24monthslog_res<- next24months_log$residuals
next24monthslog_res<- na.trim(next24monthslog_res)
```

Examine residuals and autocorrelation
```{r}
acf(next24monthslog_res, lag.max = 20)
plot.ts(next24monthslog_res)
plotForecastErrors(next24monthslog_res)
plot.forecast(next24months_log)
```

Get predictions
```{r}
next24months_log_pred<- exp(next24months_log$mean)
next24months_log_pred
```

```{r}
precip
```
Seasonal Naive
```{r}


comp_data_month_year<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), sum))
df<- data.frame(comp_data_month_year)
comp_data_df<- setNames(df, c("Month", "Year", "PRCP"))

comp_data_model_12yrs<- comp_data_df[671:815, c("Month", "Year", "PRCP")]
plot.ts(comp_data_model_12yrs$PRCP)
monthly_ts_12year<- ts(comp_data_model_12yrs$PRCP, frequency = 12)
snaivefit <- snaive(monthly_ts,h=24)

plot(snaivefit, plot.conf=FALSE,
  main="Forecasts for Monthly Precipitation")
lines(snaivefit$mean,col=3)
legend("topright", lty=1, col=3,
  legend=c("Seasonal naive method"))

snaive_res<- snaivefit$residuals
snaive_res<- na.trim(snaivefit$residuals)
plotForecastErrors(snaive_res)
snaivefit$mean
plot(snaivefit)
#testing and training sets
train.end <- time(monthly_ts)[length(monthly_ts)-24]
test.start <- time(monthly_ts)[length(monthly_ts)-24+1]
train <- window(monthly_ts,end=train.end)
test <- window(monthly_ts,start=test.start)
fitsnaive <- snaive(train, h=24)
fcsnaive <- forecast(fitsnaive,h=24)
accuracy(fcsnaive, test)
plot(fcsnaive)
lines(test, col=5)
```

ARIMA
```{r}
install.packages("tseries", repo="http://cran.us.r-project.org")
library("tseries")
comp_data_month_year<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), sum))
df<- data.frame(comp_data_month_year)
comp_data_df<- setNames(df, c("Month", "Year", "PRCP"))

comp_data_model_6yrs<- comp_data_df[743:815, c("Month", "Year", "PRCP")]
comp_data_model_6yrs$PRCPNEW<- comp_data_model_6yrs$PRCP

comp_data_model_all<- comp_data_df[, c("Month", "Year", "PRCP")]
comp_data_model_all$PRCPNEW<- comp_data_model_all$PRCP

monthly_ts<- ts(comp_data_model_all$PRCPNEW, frequency = 12, start = c(1946, 9))
monthly_ts

adf.test(monthly_ts, alternative = "stationary")


tsdisplay(diff(monthly_ts,4))
fit011 <- Arima(monthly_ts, order=c(0,1,1), seasonal=c(0,1,1))
tsdisplay(residuals(fit011))

fit013 <- Arima(monthly_ts, order=c(0,1,3), seasonal=c(0,1,1))
tsdisplay(residuals(fit013))

res011 <- residuals(fit011)
res013 <- residuals(fit013)
Box.test(res011, lag=16, fitdf=4, type="Ljung")
Box.test(res013, lag=16, fitdf=4, type="Ljung")

forecast011<- forecast(fit011, h=24)
forecast013<- forecast(fit013, h=24)
plot(forecast011)
plot(forecast013)
fitauto<- auto.arima(monthly_ts)
tsdisplay(residuals(fitauto))
resauto<- residuals(fitauto)
Box.test((resauto))
forecastauto<- forecast(fitauto, h=24)
plot(forecastauto)
fitauto
fit013
fit011
```

ARIMA with test and train datasets
```{r}
getrmse <- function(x,h,...)
{
  train.end <- time(x)[length(x)-h]
  test.start <- time(x)[length(x)-h+1]
  train <- window(x,end=train.end)
  test <- window(x,start=test.start)
  fit <- Arima(train,...)
  fc <- forecast(fit,h=h)
  return(accuracy(fc,test)[2,"RMSE"])
}

getrmse(monthly_ts,h=24,order=c(3,0,0),seasonal=c(2,1,0))
getrmse(monthly_ts,h=24,order=c(3,0,1),seasonal=c(2,1,0))
getrmse(monthly_ts,h=24,order=c(3,0,2),seasonal=c(2,1,0))
getrmse(monthly_ts,h=24,order=c(3,0,1),seasonal=c(1,1,0))
getrmse(monthly_ts,h=24,order=c(3,0,1),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(3,0,1),seasonal=c(0,1,2))
getrmse(monthly_ts,h=24,order=c(3,0,1),seasonal=c(1,1,1))
getrmse(monthly_ts,h=24,order=c(4,0,3),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(3,0,3),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(4,0,2),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(3,0,2),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(2,1,3),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(2,1,4),seasonal=c(0,1,1))
getrmse(monthly_ts,h=24,order=c(2,1,5),seasonal=c(0,1,1))

getrmse(monthly_ts,h=24,order=c(4,0,3),seasonal=c(0,1,1))

train.end <- time(monthly_ts)[length(monthly_ts)-24]
test.start <- time(monthly_ts)[length(monthly_ts)-24+1]
train <- window(monthly_ts,end=train.end)
test <- window(monthly_ts,start=test.start)

fitbest<- Arima(train, order=c(4,0,3), seasonal=c(0,1,1))
tsdisplay(residuals(fitbest))
resbest<- residuals(fitbest)
Box.test(resbest, lag=16, fitdf=4, type="Ljung")

forecastbest<- forecast(fitbest, h=24)
plot(forecastbest)
lines(test, col=5)
test
accuracy(forecastbest, test)

#Get predictions
fitbest<- Arima(monthly_ts, order=c(4,0,3), seasonal=c(0,1,1))
forecastbest<- forecast(fitbest, h=24)
plot(forecastbest)
preds<-forecastbest$mean
preds
```

ARIMA with Log Transformation
```{r}
comp_data_month_year<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), sum))
df<- data.frame(comp_data_month_year)
comp_data_df<- setNames(df, c("Month", "Year", "PRCP"))

comp_data_model_6yrs<- comp_data_df[743:815, c("Month", "Year", "PRCP")]
comp_data_model_6yrs$PRCPNEW<- comp_data_model_6yrs$PRCP

comp_data_model_all<- comp_data_df[, c("Month", "Year", "PRCP")]
comp_data_model_all$PRCPNEW<- comp_data_model_all$PRCP

monthly_ts<- ts(comp_data_model_all$PRCPNEW, frequency = 12, start = c(1946, 9))
monthly_ts

adf.test(monthly_ts, alternative = "stationary")


tsdisplay(diff(monthly_ts,4))
fit011 <- Arima(monthly_ts, order=c(0,1,1), seasonal=c(0,1,1))
tsdisplay(residuals(fit011))

fit013 <- Arima(monthly_ts, order=c(0,1,3), seasonal=c(0,1,1))
tsdisplay(residuals(fit013))

res011 <- residuals(fit011)
res013 <- residuals(fit013)
Box.test(res011, lag=16, fitdf=4, type="Ljung")
Box.test(res013, lag=16, fitdf=4, type="Ljung")

forecast011<- forecast(fit011, h=24)
forecast013<- forecast(fit013, h=24)
plot(forecast011)
plot(forecast013)
fitauto<- auto.arima(monthly_ts)
tsdisplay(residuals(fitauto))
resauto<- residuals(fitauto)
Box.test((resauto))
forecastauto<- forecast(fitauto, h=24)
plot(forecastauto)
fitauto
fit013
fit011
```

ARIMA with log transformations on test and train datasets
```{r}
comp_data_model_all<- comp_data_df[, c("Month", "Year", "PRCP")]
comp_data_model_all$PRCPNEW<- comp_data_model_all$PRCP

comp_data_month_year<- (aggregate(comp_data[,'PRCP'], list(comp_data$Month, comp_data$Year), sum))
df<- data.frame(comp_data_month_year)
df
comp_data_df<- setNames(df, c("Month", "Year", "PRCP"))
summary(comp_data_df)


comp_data_model_f<- comp_data_df[is.finite(comp_data_df)]
monthly_ts<- ts(comp_data_df$PRCP, frequency = 12, start = c(1946, 9))

monthly_ts[monthly_ts==0] <- 0.001
summary(monthly_ts)

monthly_ts_log<- log(monthly_ts)

plot(monthly_ts_log_f)
summary(monthly_ts_log_f)

getrmse <- function(x,h,...)
{
  train.end <- time(x)[length(x)-h]
  test.start <- time(x)[length(x)-h+1]
  train <- window(x,end=train.end)
  test <- window(x,start=test.start)
  fit <- Arima(train,...)
  fc <- forecast(fit,h=h)
  return(accuracy(fc,test)[2,"RMSE"])
}

getrmse(monthly_ts_log,h=24,order=c(3,0,0),seasonal=c(2,1,0))
getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(2,1,0))
getrmse(monthly_ts_log,h=24,order=c(3,0,2),seasonal=c(2,1,0))
getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(1,1,0))
getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(0,1,2))
getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(1,1,1))
getrmse(monthly_ts_log,h=24,order=c(4,0,3),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(3,0,3),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(4,0,2),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(3,0,2),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(2,1,3),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(2,1,4),seasonal=c(0,1,1))
getrmse(monthly_ts_log,h=24,order=c(2,1,5),seasonal=c(0,1,1))

getrmse(monthly_ts_log,h=24,order=c(3,0,1),seasonal=c(0,1,1))

train.end <- time(monthly_ts_log)[length(monthly_ts_log)-24]
test.start <- time(monthly_ts_log)[length(monthly_ts_log)-24+1]
train <- window(monthly_ts_log,end=train.end)
test <- window(monthly_ts_log,start=test.start)

fitbestlog<- Arima(train, order=c(3,0,1),seasonal=c(0,1,1))
tsdisplay(residuals(fitbestlog))
resbestlog<- residuals(fitbestlog)
Box.test(resbestlog, lag=16, fitdf=4, type="Ljung")

forecastbestlog<- forecast(fitbestlog, h=24)
preds<- forecastbestlog$mean
lines(train[700:815])
accuracy(forecastbestlog, test)

#Get predictions
fitbestlog<- Arima(monthly_ts_log_f, order=c(3,0,1),seasonal=c(0,1,1))
forecastbestlog<- forecast(fitbestlog, h=24)
forecastbestlog$mean
plot(forecastbestlog)
preds<- exp(forecastbestlog$mean)

plot(monthly_ts)
lines(preds, col=4)


```
